Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex
Switch to reference for URL

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.
Recent additions using Zarr include the application of deep learning
models to genomic sequence data~\citep{klie2023predictive} 
and a browser for genetic variation data~\citep{konig2023divbrowse}.
Zarr has in fact been in use in genomics for many years, and was originally 
developed to store data from the \textit{Anopheles gambiae} 1000 Genomes Project~\citep{anopheles2017genetic}.
A successor project, the MalariaGEN Vector Observatory,
currently uses Zarr to store data from whole-genome sequencing of 23,000 \textit{Anopheles} mosquitoes from 31 African countries.\footnote{\url{https://www.malariagen.net/malaria-vector-genome-observatory/}}
These data are hosted in Google Cloud Storage and can be analysed interactively
using free cloud computing services like Google Colab,
democratising the use of data by scientists in malaria-endemic countries
where access to local computing infrastructure and sufficient network bandwidth to download large datasets may be limited.

>>>>

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.
Recent additions using Zarr include the application of deep learning
models to genomic sequence data~\citep{klie2023predictive} 
and a browser for genetic variation data~\citep{konig2023divbrowse}.
Zarr has in fact been in use in genomics for many years, and was originally 
developed to store data from the \textit{Anopheles gambiae} 
1000 Genomes Project~\citep{anopheles2017genetic}.
A successor project, the MalariaGEN Vector Observatory,
currently uses Zarr to store data from whole-genome sequencing 
of 23,000 \textit{Anopheles} mosquitoes from 31 African 
countries~\cite{mvgo2024}.
The data is hosted in Google Cloud Storage and can be analysed interactively
using free cloud computing services like Google Colab,
democratising the use of data by scientists in malaria-endemic countries
where access to local computing infrastructure and sufficient network 
bandwidth to download large datasets may be limited.
